---
title: "Diagnosing a DEMATEL system"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnosing a DEMATEL system}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

Two DEMATEL models can have the same number of factors, the same rating scale
and the same panel size, and still require opposite interventions. Which one you
are holding is a property of the matrix's spectrum rather than of its subject
matter, so **you cannot look up your intervention logic, you have to diagnose
it.**

This vignette works through that diagnosis on the two published matrices the
package ships.

```{r setup}
library(spectralDEMATEL)

fefo <- worked_matrices$fefo_stock_control          # 13 factors
resl <- worked_matrices$resilience_capabilities     # 15 factors
```

## Is this matrix in scope?

Before any number means anything, the matrix has to satisfy the conditions the
diagnostics rest on. Every one of them comes back as data — nothing here stops,
warns or prints.

```{r}
ck <- assumption_checks(fefo)
ck[ck$verdict != "pass", c("check", "verdict", "value")]
```

One warning, and it is worth having. This matrix has a non-zero diagonal:
thirteen factors carry a self-influence. That is admissible and everything is
computed as usual, but self-influence inflates coupling, so the value is not
comparable with a study that set its diagonal to zero.

The second matrix is a harder case.

```{r}
ck2 <- assumption_checks(resl)
ck2[ck2$verdict != "pass", c("check", "verdict", "value")]
```

Strong connectivity fails. A binary verdict would stop there; the point of
returning the checks as data is that they can say **which factors**:

```{r}
stranded <- ck2$factors[ck2$check == "strong_connectivity"][[1]]
stranded
```

The chain of cause is visible in the matrix itself. Row 14 is blank, so factor
14 dispatches nothing. Column 7 is blank, so factor 7 receives nothing. Factor
8's only source is 7, and factor 4's only sources are 7 and 8 — so neither can
be reached from the rest of the system either.

```{r}
which(rowSums(resl) == 0)   # dispatches nothing
which(colSums(resl) == 0)   # receives nothing
```

That is usually a coding slip in one row rather than a property of the system,
which is exactly why naming the factors matters: a user is told where to look.

The engine computes the diagnostics anyway. This is deliberate — the published
study reports them, and an engine that refused would make the check
unimplementable. But the numbers carry the verdict with them, and here they
carry a visible mark of it:

```{r}
d_resl <- spectral_diagnostics(resl)

round(d_resl$entry_points[stranded], 4)   # the top-ranked entry point is here
round(d_resl$accumulation[stranded], 4)   # yet nothing accumulates at them
```

The highest-ranked entry point in the whole system is one of the four factors
cut off from it, and the left eigenvector puts exactly zero weight on three of
them. For a strongly connected matrix the entry profile is unique and strictly
positive; when A2 fails it is neither, and a ranking read off it is describing
something other than the system as a whole.

### Why that is worth the trouble

Here is the mechanism on a matrix small enough to read. Five factors, densely
and unevenly connected:

```{r}
core <- matrix(c(0, 4, 3, 1, 1,
                 4, 0, 3, 1, 1,
                 2, 2, 0, 1, 1,
                 1, 1, 1, 0, 1,
                 1, 1, 1, 1, 0), 5, 5, byrow = TRUE)
storage.mode(core) <- "double"

d_core <- spectral_diagnostics(core)
c(irreducible = is_irreducible(core),
  coupling = round(d_core$mu_max, 3), hierarchy = round(d_core$hierarchy_sd, 3))
```

Coupling above 0.5 and hierarchy above 0.10: strongly coupled, with influence
entering at a few factors.

Now add a sixth factor that dispatches heavily to all five and receives nothing
— an observer, a reporting layer, a context variable someone added late.

```{r}
observed <- rbind(cbind(core, 0), 0)
observed[6, 1:5] <- 4

d_obs <- spectral_diagnostics(observed)
c(irreducible = is_irreducible(observed),
  coupling = round(d_obs$mu_max, 3), hierarchy = round(d_obs$hierarchy_sd, 3))
```

Coupling has fallen from `r round(d_core$mu_max, 2)` to
`r round(d_obs$mu_max, 2)`, crossing the 0.5 threshold, so the system reads as
dampened where it was amplified. Nothing about the five original factors
changed. What changed is the normalising constant: `dematel()` divides by the
larger of the maximum row and column total, and factor 6's row is now the
largest, so the whole normalised matrix shrank.

```{r}
c(s_core = dematel(core)$s, s_observed = dematel(observed)$s)
```

The check catches this. The verdict is not bureaucratic.

## What kind of system is this?

```{r}
d_fefo <- spectral_diagnostics(fefo)

data.frame(
  row.names  = c("FEFO stock control", "Resilience capabilities"),
  factors    = c(d_fefo$n, d_resl$n),
  coupling   = c(d_fefo$mu_max, d_resl$mu_max),
  multiplier = c(d_fefo$multiplier, d_resl$multiplier),
  indirect   = c(d_fefo$indirect_dominant, d_resl$indirect_dominant),
  dominance  = c(d_fefo$dominance, d_resl$dominance)
)
```

The two are opposites. FEFO stock control is strongly coupled: total effects run
`r round(d_fefo$multiplier, 1)` times the direct ones, so indirect propagation
dominates and an intervention has to be appraised on total effects rather than
on the ratings as entered. Its mode dominance of `r round(d_fefo$dominance, 3)`
says one propagation pattern accounts for essentially all of the behaviour.

Resilience capabilities barely amplifies at all: a multiplier of
`r round(d_resl$multiplier, 2)` means indirect effects are a third of the direct
ones. Direct effects are most of the story, and the ratings can be read close to
face value.

Coupling, the dominant eigenvalue and the multiplier are one quantity on three
scales, which the package asserts rather than assumes:

```{r}
c(from_mu     = 1 / (1 - d_fefo$mu_max),
  from_lambda = 1 + d_fefo$lambda_max,
  reported    = d_fefo$multiplier)
```

## Reading hierarchy in the right direction

Hierarchy has three readings and **they do not run the same way**.

```{r}
data.frame(
  row.names   = c("FEFO stock control", "Resilience capabilities"),
  sd          = c(d_fefo$hierarchy_sd,   d_resl$hierarchy_sd),
  gini        = c(d_fefo$hierarchy_gini, d_resl$hierarchy_gini),
  pr          = c(d_fefo$hierarchy_pr,   d_resl$hierarchy_pr)
)
```

`hierarchy_sd` and `hierarchy_gini` are **high** when influence enters at a few
factors. `hierarchy_pr` is **low**. Resilience capabilities is the more
concentrated of the two on all three readings, which you can only see if you
know that the third column runs backwards.

Anything that displays one of these must say which reading it used and which way
it runs. This project has inverted the direction once already, in code that had
been reviewed, and the mistake produced entirely plausible output.
`hierarchy_sd` is the reading behind every published corpus-level result.

## Where do I intervene?

The standard DEMATEL deliverable is prominence, the sum of a factor's row and
column totals in the total-relation matrix. It adds two different roles
together: how much a factor dispatches, and how much it absorbs. The left and
right eigenvectors separate them.

```{r}
T_fefo <- dematel(fefo)$T
prominence <- rowSums(T_fefo) + colSums(T_fefo)

ranks <- data.frame(
  factor       = seq_len(d_fefo$n),
  prominence   = rank(-prominence),
  entry        = rank(-d_fefo$entry_points),
  accumulation = rank(-d_fefo$accumulation)
)
head(ranks[order(ranks$entry), ], 5)
```

The three rankings disagree, and the disagreements are the interesting part.

Factor 4 is the **second strongest entry point and only seventh by prominence**.
Factor 3 is third by entry and fifth by prominence. Both inject more influence
than the standard deliverable credits them with.

Factor 13 runs the other way: **second by prominence, first by accumulation,
only fourth by entry**. Prominence adds dispatch to absorption, so a factor that
absorbs heavily while dispatching moderately can outrank the factor actually
driving the system. Pressure applied at 13 lands where effects already
accumulate, which is close to the opposite of intervening.

Only factor 5 tops all three, which is why a single blended score is
persuasive right up to the point where it is wrong.

### Which relationship, not which factor

`sensitivity_matrix()` gives the derivative of the dominant eigenvalue in every
off-diagonal entry, so the question moves from *which factor matters* to *which
relationship, if strengthened or weakened, moves the system most.*

```{r}
s <- sensitivity_matrix(fefo)
strongest <- which(s$total == max(s$total, na.rm = TRUE), arr.ind = TRUE)
strongest
```

**Never read that ranking without the condition number.**

```{r}
d_fefo$ev_condition
```

`ev_condition` is at least 1 by construction and bounds how far a first-order
estimate can be trusted. At `r round(d_fefo$ev_condition, 2)` these derivatives
can be read at face value; a value in the tens means the derivative is locally
uninformative, and `assumption_checks()` raises a warning verdict past 5.

The derivative is signed, and some entries are negative:

```{r}
sum(s$total < 0, na.rm = TRUE)
```

Strengthening one of those *lowers* coupling — the lever for a system someone
wants to calm rather than drive, which is the usual situation in risk work. The
reason is not mysterious, and the function separates it out:

```{r}
neg <- which(s$total < 0, arr.ind = TRUE)[1, ]
c(local         = s$local[neg[1], neg[2]],
  normalization = s$normalization[neg[1], neg[2]],
  total         = s$total[neg[1], neg[2]])
```

The local effect is always an amplification. The normalisation effect is never
one, and it is non-zero only in the row or column that currently sets the
normalising constant. Where the second outweighs the first, raising an influence
rating lowers the dominant eigenvalue. Shown as a single total, that reads as a
bug; shown as two components, it is a mechanism.

## How much should I trust that reading?

A diagnosis is worth less if the same numbers would arise from the rating
distribution alone. `surrogate_ensemble()` shuffles the off-diagonal entries,
holding the number of factors, the density and the exact multiset of values
fixed, and recomputes.

```{r}
ens <- surrogate_ensemble(fefo, B = 200, seed = 42)

c(observed = d_fefo$hierarchy_sd,
  surrogate_median = median(ens$hierarchy_sd),
  share_at_least_observed = mean(ens$hierarchy_sd >= d_fefo$hierarchy_sd))
```

Read that carefully, because it does not say what a null test usually says. The
observed hierarchy is **lower** than the surrogate median: shuffling this
matrix's own ratings at random produces a *more* concentrated entry profile than
the matrix actually has, about
`r round(100 * mean(ens$hierarchy_sd >= d_fefo$hierarchy_sd))`% of the time. The
diffuseness of this system is not an artefact of its rating distribution — it is
structure the analyst built, and it survives the comparison.

The same ensemble answers the question for coupling and dominance, and there the
verdict is sharper still:

```{r}
data.frame(
  row.names = c("coupling", "dominance", "hierarchy_sd"),
  observed  = c(d_fefo$mu_max, d_fefo$dominance, d_fefo$hierarchy_sd),
  surrogate_min = c(min(ens$mu_max), min(ens$dominance), min(ens$hierarchy_sd)),
  surrogate_max = c(max(ens$mu_max), max(ens$dominance), max(ens$hierarchy_sd))
)
```

Coupling sits **above every one of the 200 surrogates** and mode dominance
**below every one of them**. Shuffling this matrix's ratings cannot reproduce
either: the system is more strongly coupled, and more completely governed by a
single propagation mode, than its own entry distribution can account for. Only
hierarchy falls inside the ensemble, and on the low side.

That is the comparison the source paper could not run across its corpus, because
the precomputed spectral file holds no raw matrices. A user of this package has
their own matrix, so they can.

The seed is exposed so a figure can be reproduced. This is the only function
here that is not instant — each draw is an eigendecomposition plus a
connectivity check — so run it on demand rather than eagerly, and lower `B` if a
user is waiting.

## Showing your work

Every diagnosis records the version of the engine that produced it.

```{r}
d_fefo$engine_version
```

A metric definition has moved in this project's history before. When one moves
again, a diagnosis recorded with its version stays interpretable rather than
silently meaning something new. The checks travel in the same object for the
same reason: numbers whose validity depends on something the caller was told
separately are numbers whose validity will eventually be lost.

```{r}
names(d_fefo)
```
